\chapter*{LIST OF SYMBOLS AND ABBREVIATIONS}
\addcontentsline{toc}{chapter}{LIST OF SYMBOLS AND ABBREVIATIONS}

\section*{List of Abbreviations}
\begin{tabularx}{\textwidth}{@{}p{0.23\textwidth}X@{}}
\textbf{ASR} & Automatic Speech Recognition \\
\textbf{BLEU} & Bilingual Evaluation Understudy score, commonly used in sequence generation quality assessment \\
\textbf{BiLSTM} & Bidirectional Long Short-Term Memory network \\
\textbf{CNN} & Convolutional Neural Network \\
\textbf{CSLR} & Continuous Sign Language Recognition \\
\textbf{CTC} & Connectionist Temporal Classification \\
\textbf{DGS} & Deutsche Gebardensprache (German Sign Language) \\
\textbf{FPS} & Frames Per Second \\
\textbf{GUI} & Graphical User Interface \\
\textbf{HMM} & Hidden Markov Model \\
\textbf{I3D} & Inflated 3D ConvNet for spatiotemporal feature extraction \\
\textbf{PE} & Positional Encoding \\
\textbf{ReLU} & Rectified Linear Unit activation function \\
\textbf{ResNet} & Residual Neural Network \\
\textbf{RWTH-PHOENIX} & RWTH-PHOENIX-Weather 2014 benchmark dataset \\
\textbf{SLR} & Sign Language Recognition \\
\textbf{VAC} & Visual Alignment Constraint \\
\textbf{VRAM} & Video Random Access Memory on GPU hardware \\
\textbf{WER} & Word Error Rate \\
\end{tabularx}

\vspace{0.75cm}

\section*{List of Symbols}
\begin{tabularx}{\textwidth}{@{}p{0.23\textwidth}X@{}}
\textbf{$\lambda_{CE}$} & Weight for cross-entropy decoder loss in the joint objective \\
\textbf{$\lambda_{CTC}$} & Weight for CTC loss in the joint objective \\
\textbf{$d_{model}$} & Transformer hidden representation dimension \\
\textbf{$Q, K, V$} & Query, key, and value projections in self-attention \\
\textbf{$T$} & Sequence length in frames or timesteps \\
\textbf{$h$} & Number of attention heads \\
\textbf{$D$} & Feature dimension of latent representation \\
\textbf{$\hat{Y}$} & Predicted gloss sequence \\
\textbf{$Y$} & Ground-truth gloss sequence \\
\textbf{$\mathcal{L}_{CTC}$} & Connectionist Temporal Classification loss term \\
\textbf{$\mathcal{L}_{CE}$} & Cross-entropy loss term for autoregressive decoding \\
\textbf{$\mathcal{L}_{hybrid}$} & Total hybrid training objective combining CTC and CE terms \\
\end{tabularx}

\vspace{0.75cm}

\section*{Key Technical Terms}
\begin{tabularx}{\textwidth}{@{}p{0.30\textwidth}X@{}}
\textbf{Co-articulation} & Temporal blending between neighboring signs that makes exact boundary localization difficult. \\
\textbf{Deletion Error} & A reference gloss omitted in prediction; typically increases when alignment confidence is weak. \\
\textbf{Insertion Error} & An extra predicted gloss not present in the reference sequence. \\
\textbf{Substitution Error} & A reference gloss replaced by another gloss in prediction due to visual or temporal confusion. \\
\textbf{Alignment Drift} & Shift of predicted sign boundaries over time, causing duplicated or missed signs near transitions. \\
\textbf{Teacher Forcing} & Training strategy where ground-truth previous tokens are provided to the decoder at each step. \\
\textbf{Beam Search} & Heuristic decoding strategy that keeps top candidate sequences rather than only the best single path. \\
\textbf{Generalization} & Model ability to maintain performance on unseen signer sequences and conditions. \\
\textbf{Gloss} & A textual token representing the canonical label of a sign unit used as the target sequence in CSLR. \\
\textbf{HamNoSys} & Hamburg Notation System, a formal sign-language notation scheme used to represent handshape, location, and movement. \\
\end{tabularx}
